Predicting instabilities of a tuneable ring laser with an iterative map model

### 2020-11-29 Papers/Journal/Tuneable_Lasers.tex
Fixed passive voice

Several effects arise within the cavity due to the interplay of dispersion, modulation, and the nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, oktem2010, shao2019, woodward2018}. The two effects of most interest for this paper are wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to be redshifted, and the trailing edge to be blueshifted (in the anomalous dispersion regime, the effect is reversed) to the point that a shock develops~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The frequency shift causes the pulse to become more rectangular in the frequency domain, with a linear chirp over most of the pulse, saturating at the leading and trailing edge. In optics, wave breaking manifests from self-phase modulation (SPM), which is a direct effect of the Kerr nonlinearity, causing the pulse to interfere with itself. New frequencies are then generated such that a rectangular profile is induced~\cite{agrawal2013, woodward2018}. Modulation instability---which typically arises in the anomalous dispersion regime---is the other effect that will be of interest. However, in the presence of multiple frequencies, modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation and four-wave mixing~\cite{agrawal1987, agrawal2013, haelterman1992}. Modulation instability causes a pulse to degrade into much shorter, less intense pulses, and can cause a laser to lose coherence~\cite{agrawal1987, coen1997, haelterman1992}. Outside of optics, wave breaking and modulation instability occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. This gives rise to the need of understanding the rich landscape of the parameter space, and determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

@@

The GNLSE, \eqref{eq:nlse}, has applications in nonlinear optics and fibre optic communications, however, it is regularly generalized even further to include a modulation term. The resulting equation is referred to as the master equation of mode-locking (see \cite{haus1975, haus1984, haus2000, tamura1996, usechak2005}), or sometimes the Ginzburg--Landau equation. Due to the nonlinear term in \eqref{eq:nlse} and the often non-trivial form of $g(A)$, as well as the possible addition of a modulation term, no analytic solution is known to the master equation. See Haus~\cite{haus2000} for a comprehensive description and history of the theory of mode-locked lasers. While, the GNLSE well represents the dynamics in, say, a single optical fibre, it is not entirely representative of the underlying physics within a laser cavity such as Fig. \ref{fig:cavity}. It is assumed in \eqref{eq:nlse} that each process affects the pulse continuously within the cavity. Therefore, the GNLSE, and its generalizations, can be thought of as an `average' model for lasers---this assumption breaks down if the pulse undergoes large changes during any one round trip. Often, this is the case due to the specialized, discrete components.

@@

Iterative map methods have been used for optics since 1955 when Cutler~\cite{cutler1955} analyzed a microwave regenerative pulse generator. Siegman and Kuizenga~\cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman1969} adapted Cutler's method for both AM and FM mode-locked lasers in 1969, and Martinez \emph{et al.}~\cite{martinez1984, martinez1985} incorporated the nonlinearity while modelling passively mode-locked lasers. This method was used for dispersion-tuned actively mode-locked lasers recently by Burgoyne \emph{et al.}~\cite{burgoyne2014}. Each of these models used a prescribed transfer function to describe the effect each component has on the pulse. While the transfer functions are chosen suitably to have the effect observed in the laboratory, the transfer functions are phenomenologically chosen, and not necessarily obtained directly from the underlying physics. Furthermore, these methods have not included the effect of every component, lacking the imperative interplay of dispersion, modulation, and nonlinearity---hindering their models to exhibit modulation instability.

@@

\subsection{Non-Dimensionalization}
Non-dimensionalization is a technique used in mathematical modelling to better understand the relative importance of different processes within a system. In essence, we scale all variables by typical values, based on the geometry and boundary conditions (see, for example,~\cite{howison2005}). Doing so `factors' the dimensional units out of the problem, leaving only dimensionless equations with dimensionless parameters. The relative magnitudes of the dimensionless parameters characterize the dominant dynamics---like how fluid flow can be described by the Reynolds number. Each process in the laser can be better understood by re-scaling the time, energy, and amplitude by the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}

@@

Now that we have the algebraic effect of each section of the cavity, we are ready to take the composition of the maps to give the effect of one round trip of the cavity. Unlike the models described in the beginning of this section, the permutation of our components is important as the maps \eqref{eq:effects} do not necessarily commute with each other. The geometry of a particular laser will dictate the order of the process maps. We assume the laser is constructed as in Fig.~\ref{fig:cavity}, and provide the following justifications. The effect of the nonlinearity will be maximal after the pulse passes through the gain fibre---this is the region where the pulse is most energetic. Thus, we take the nonlinearity component to follow the gain. Similarly, we would want the loss component to directly follow the nonlinearity in an attempt to minimize the effect the nonlinearity has on the pulse. This leaves the dispersion and modulation components. For now, let us assume the CFBG precedes the modulator. Finally, we assume the loss component is our starting point since this corresponds to the optical coupler, and thus, allows us to more easily compare to experiments. More precisely, let
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A)))))
	\label{eq:order}
\end{align}
denote one complete round trip of the cavity. Since a ring laser is cyclic in nature, we use the result of one round trip as the input pulse for the next round trip---restarting the process. We continually repeat this procedure until the envelope and chirp of the pulse reach equilibrium, or the pulse succumbs to modulation instability. We are uninterested in the phase change, as it is not modified by \eqref{eq:effects} in a meaningful way, and since the phase cannot easily be measured experimentally. An example of the evolution of the envelope during one round trip of the laser cavity can be found in Fig.~\ref{fig:cavityevo}.

@@

\subsection{Linear Solution}
By neglecting the effect of the nonlinearity, that is, $b = 0$, a solution can be found analytically. We proceed by assuming a fixed point solution will take the form of a linearly chirped Gaussian. The reasons for this assumption are that the solution to the models presented in~\cite{cutler1955, siegman1969, kuizenga1970a, martinez1984, martinez1985} were Gaussian, the equilibrium shape will be correlated to the modulation function, and a Gaussian is a fixed point of the Fourier transform. Furthermore, this form is chosen because it resembles the envelope and linear chirp expected from the literature~\cite{burgoyne2014, haus1975, haus1996, haus2000, usechak2005}.

@@

Consider the pulse
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \exp(i \phi_0),
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the phase (Fig.~\ref{fig:samplegauss}). By computing \eqref{eq:A0} after one round trip, and solving $\mathcal{L}(A) = A \textrm{e}^{i \phi}$, a tedious manipulation yields the conditions
\begin{align}
	\frac{\sigma^4}{1 - \sigma^2} &= (\sigma^2 + 2 C s^2)^2 + 4 s^4, \label{eq:sigrelation} \\
	\frac{C}{1 - \sigma^2} &= C + 2 \frac{s^2}{\sigma^2} (1 + C^2), \label{eq:chirprelation} \\
	1 &= \frac{W(a E_g \textrm{e}^{E_g})}{E_g} h^2 (1 - \sigma^2)^{1/2} \label{eq:energyrelation}
\end{align}
for \eqref{eq:A0} to be in equilibrium. Manipulating \eqref{eq:sigrelation} and \eqref{eq:chirprelation} we are able to eliminate the chirp, and find the expression
\begin{align}
	\sigma^8 + 4 s^4 \sigma^6 - 20 s^4 \sigma^4 + 32 s^4 \sigma^2 - 16 s^4 = 0,
	\label{eq:var}
\end{align}
which is in terms of $\sigma^2$ only. As \eqref{eq:var} is a quartic in $\sigma^2$, it has an analytic solution, namely,
\begin{equation}
	\begin{split}
		\sigma^2 = \sqrt{2} s \left( s^6 + 3s^2 + \sqrt{4 + s^4} \left( 1 + s^4 \right) \right)^{1/2} & \\
		- s^4 - s^2 & \sqrt{4 + s^4}.
		\label{eq:equilvar}
	\end{split}
\end{equation}
Additionally, the chirp is
\begin{align}
	C = \frac{\sigma^4 - \left( \sigma^8 - 16 s^4 \left( 1 - \sigma^2 \right)^2 \right)^{1/2}}{4 s^2 \left( 1 - \sigma^2 \right)},
	\label{eq:chirp}
\end{align}
in terms of $s$ and $\sigma^2$. Moreover, we can obtain the energy of the pulse entering the gain fibre, $E_g$, from \eqref{eq:energyrelation}, and thus, the equilibrium energy at the optical coupler. Doing so gives
\begin{align}
	E &= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{1 - h^2 \left( 1 - \sigma^2 \right)^{1/2}}.
	\label{eq:analenergy}
\end{align}
By equating the energy of the pulse in \eqref{eq:A0} and \eqref{eq:analenergy}, we find the peak power to be
\begin{align}
	P &= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{\sqrt{\pi} \sigma \left( 1 - h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}.
	\label{eq:analpower}
\end{align}

The nature of \eqref{eq:var}--\eqref{eq:analpower} makes the expressions difficult to work with. However, by asymptotically expanding \eqref{eq:var}--\eqref{eq:analpower} in terms of the dispersion parameter, $s$, we find the simpler relations
\begin{align}
	\sigma^2 &\sim 2s(1 - s) + \bigO{s^3}, \\
	C &\sim 1 - s + \frac{1}{2}s^2 + \bigO{s^3}, \\
	E &\sim \frac{\log (a h^2)}{1 - h^2} - \frac{1}{1 - h^2} \left( 1 + \frac{h^2 \log(a h^2)}{1 - h^2}  \right) s + \bigO{s^2}, \label{eq:maxenergy} \\
	P &\sim \frac{\log(ah^2)}{\sqrt{2 \pi}(1 - h^2)} s^{-1/2} + \bigO{s^{1/2}},
\end{align}
in the case $s \rightarrow 0$. And,
\begin{align}
	\sigma^2 &\sim 1 - \frac{1}{4}s^{-4} + \bigO{s^{-8}}, \label{eq:siginfty}\\
	C &\sim \frac{1}{2}s^{-2} + \bigO{s^{-6}},
\end{align}
in the case $s \rightarrow \infty$. We have no asymptotic expansion of the energy or peak power as $s \rightarrow \infty$ since in this limit too much energy is lost due to dispersion and modulation. The gain is not powerful enough to balance the energy lost. The argument of the logarithm in \eqref{eq:analenergy} gives us a condition for the pulse to be sustainable. Expanding the requirement $a h^2 (1 - \sigma^2)^{1/2} > 1$, from \eqref{eq:analenergy}, using \eqref{eq:siginfty} (for simple approximation), we find 
\begin{align}
	s^* = \left( \frac{a h^2}{2} \right)^{1/2}
\end{align}
is the maximum value of $s$ to facilitate a sustainable pulse.

@@

In Fig.~\ref{fig:chirpft} we show the chirp and Fourier transform of the pulse \eqref{eq:nlA0} after it has reached equilibrium for the nominal parameter values given in \eqref{eq:ndparam}, and two additional $b$ values. In this nonlinear case, we find that the equilibrium pulse is no longer Gaussian---as evidenced by the Fourier transforms. Notice that compared to the linear case, higher frequency modes are introduced due to the nonlinearity, giving a bi-modal distribution. Moreover, we find the pulse is linearly chirped near the peak, but, in the tails, $|T| > 1$, the chirp begins to saturate---consistent with the experimental results~\cite{chen2008, rothenberg1989b, tomlinson1985}. As the nonlinearity parameter, $b$, is increased, the chirp increases more sharply across the pulse, and saturates at a larger value. Also, the two peaks in the Fourier transform become more pronounced and further apart, and the Fourier transform of the pulse becomes more rectangular. However, after a critical point, SPM plays a more substantial role, leading to modulation instability. This phenomenon is highlighted in Fig.~\ref{fig:breakevo}. By decreasing the dispersion from to Fig.~\ref{fig:chirpft}, we find the pulse begins to `breathe'. During the first two dozen round trips of the cavity, the SPM compounds and becomes detrimental to the pulse. This in turn induces modulation instability, and degrades the pulse, until the pulse is no longer stable or sustainable. On the other hand, having a more moderate value of $b$, the pulse is able to equilibrate even with a less favourable, random, initial pulse (Fig.~\ref{fig:convevo}), as in~\cite{meng2020}. The pulse is able to shed the initial left lobe ($T \approx -1 / 2$) because of dispersion and modulation, and a central lobe ($T \approx 0$) is formed and is able to grow due to the gain medium. This then allows the pulse to come to equilibrium very quickly. Furthermore, since the intensity of the initial pulse is relatively small, the effect of SPM is negligible during the first few round trips, allowing the laser to select preferable modes. The preferable modes then get amplified, stabilizing the pulse. It is clear the initial shape of the pulse is much less important than the initial energy of the pulse, $E_0$, or the strength of the nonlinearity, $b$.

@@

We now turn our attention to a more quantitative analysis, and characterize the stability of a laser by exploring the parameter space. We focus on the $s$--$b$ plane in particular, since $a$ and $h$ effect only the amplitude of the pulse, whereas the interplay of dispersion and nonlinearity gives rise to the instabilities. Additionally, as mentioned, a conservative initial energy will effect the dynamics much less than a more intense pulse since the nonlinearity will be negligible while initially coming to equilibrium. To identify the stability of a pulse, we examine the relative change in the pulse's envelope between consecutive round trips of the cavity. We compute this error as
\begin{align}
	\Delta = \frac{\| |\mathcal{L}^n(A_0)| - |\mathcal{L}^{n-1}(A_0)| \|_2}{\| \mathcal{L}^{n-1}(A_0) \|_2},
	\label{eq:error}
\end{align}
where
\begin{align}
	\| f \|_2^2 \coloneqq \int_{-\infty}^\infty |f|^2 \, \df T,
\end{align}
and $n$ is chosen to be sufficiently large to guarantee the pulse reaches equilibrium or deteriorates due to modulation instability---we use $n = 500$ in our experiments. We take the modulus of the pulse since we are only interested in the evolution of the envelope because of the auto-correlation methods used for experimental measurements. We plot the error in Fig.~\ref{fig:error}. The parameter space is split into two distinct regions, divided by an incredibly sharp and complicated boundary. In the upper-left region, the error is $\bigO{1}$; here, the nonlinearity is too strong, and the SPM induces modulation instability. In the lower-right region we find the opposite behaviour---the pulse reaches an equilibrium state and is stable. The increase in dispersion, coupled with modulation, allows the laser to balance the nonlinear effects. The error in this region is on the order of machine precision.

@@

Also of interest is the energy of the pulse at equilibrium, which we compute numerically by \eqref{eq:energy}. Given our ordering of the components, \eqref{eq:order}, the energy of the pulse is immediately before it passes through the optical coupler. The energy of each pulse emitted by the laser is $(1 - h^2) E$, and $h^2 E$ remains in the cavity. We plot the energy in the $s$--$b$ plane in Fig.~\ref{fig:energy}. Unsurprisingly, we find the same sharp boundary as we saw in Fig.~\ref{fig:error}. In the unstable region (upper-left) the energy is relatively small. Since the pulse is not sustainable in this region it is not able to foster prominent modes, thus, the intensity is low across the entire pulse. In the stable region (lower-right) we find that the energy smoothly decays as both $b$ and $s$ increase, with the contours being hyperbolic-esque. As $s$ increases, more energy is lost due to modulation since dispersion increases the length of the tails of the pulse. Similarly, as $b$ increases, the additional frequencies generated by SPM are suppressed by our choice of modulation function. We find that a more energetic laser requires weaker dispersion, however, the nonlinearity must be correspondingly low, otherwise modulation instability will destroy the pulse. Moreover, the maximum energy, which occurs in the limit that both $s$ and $b$ go to zero, is 2.5487. With our choice of parameters, \eqref{eq:maxenergy} predicts an equilibrium energy of 2.5535 at the origin, confirming the usefulness of the approximation.

@@

\section{Conclusion}
\label{sec:conclusion}
Expanding upon the ideas originally proposed by Cutler~\cite{cutler1955}, and Kuizenga and Siegman~\cite{kuizenga1970, kuizenga1970a, siegman1969}, we developed a nonlinear iterative map model for tuneable ring lasers. We recovered linearly chirped Gaussian solutions in the case of a Gaussian modulation function when omitting the nonlinearity. Moreover, the solution to the linear model can be extended to any modulation function using results of~\cite{calcaterra2008a}. In contrast, with the inclusion of the nonlinearity, we were able to recover wave breaking and modulation instability, and found a sharp boundary of stability. This phenomenon has been demonstrated in a laboratory setting~\cite{agrawal2013, anderson1992, finot2008, rothenberg1989b, tomlinson1985}, but, has proven difficult to predict with simple mathematical models~\cite{meng2020}. We've shown that a simple iterative map model is able to reproduce complicated instabilities arising naturally from the interplay between dispersion, modulation, and nonlinear effects. Our model can easily be adapted to specific laser geometries, and to more complicated models of the components. For example, incorporating a frequency dependence in the gain function, or using a saturable absorber instead of a Gaussian modulator.

>>>>

Several effects arise within the cavity due to the interplay of dispersion, modulation, and the nonlinearity~\cite{bohun2015, coen1997, lapre2019, meng2020, oktem2010, shao2019, woodward2018}. The two effects of most interest for this paper are wave breaking, and modulation instability. Wave breaking causes the leading edge of a pulse to redshift, and the trailing edge to blueshift (the effect reverses in the anomalous dispersion regime) to the point that a shock develops~\cite{anderson1992, rothenberg1989a, rothenberg1989b, tomlinson1984, tomlinson1985}. The frequency shift causes the pulse to become more rectangular in the frequency domain, with a linear chirp over most of the pulse, saturating at the leading and trailing edge. In optics, wave breaking manifests from self-phase modulation (SPM), which is a direct effect of the Kerr nonlinearity, causing the pulse to interfere with itself. New frequencies are then generated such that a rectangular profile is induced~\cite{agrawal2013, woodward2018}. Modulation instability---which typically arises in the anomalous dispersion regime---is the other effect that will be of interest. However, in the presence of multiple frequencies, modulation instability can emerge in the normal dispersion regime as well through cross-phase modulation and four-wave mixing~\cite{agrawal1987, agrawal2013, haelterman1992}. Modulation instability causes a pulse to degrade into much shorter, less intense pulses, and can cause a laser to lose coherence~\cite{agrawal1987, coen1997, haelterman1992}. Outside of optics, wave breaking and modulation instability occur in other areas with nonlinear waves and dispersive media such as plasma physics, heat propagation, transmission lines, and fluid dynamics~\cite{coen1997, rothenberg1989b}. Within a ring laser cavity wave breaking and modulation instability can become parasitic leading to an unstable and unsustainable pulse. This gives rise to the need of understanding the rich landscape of the parameter space, and determining design principles to guarantee the ring laser is stable and sustainable~\cite{bohun2015, burgoyneemail, finot2008, lapre2019, woodward2018}.

@@

The GNLSE, \eqref{eq:nlse}, has applications in nonlinear optics and fibre optic communications, however, it is regularly generalized even further to include a modulation term. The resulting equation is referred to as the master equation of mode-locking (see \cite{haus1975, haus1984, haus2000, tamura1996, usechak2005}), or sometimes the Ginzburg--Landau equation. Due to the nonlinear term in \eqref{eq:nlse} and the often non-trivial form of $g(A)$, as well as the possible addition of a modulation term, no analytic solution is known to the master equation. See Haus~\cite{haus2000} for a comprehensive description and history of the theory of mode-locked lasers. While, the GNLSE well represents the dynamics in, say, a single optical fibre, it is not entirely representative of the underlying physics within a laser cavity such as Fig. \ref{fig:cavity}. An assumption of \eqref{eq:nlse} is that each process affects the pulse continuously within the cavity. Therefore, the GNLSE, and its generalizations, can be thought of as an `average' model for lasers---this assumption breaks down if the pulse undergoes large changes during any one round trip. Often, this is the case due to the specialized, discrete components.

@@

Iterative map methods have been used for optics since 1955 when Cutler~\cite{cutler1955} analyzed a microwave regenerative pulse generator. Siegman and Kuizenga~\cite{kuizenga1970a, kuizenga1970b, kuizenga1970, siegman1969} adapted Cutler's method for both AM and FM mode-locked lasers in 1969, and Martinez \emph{et al.}~\cite{martinez1984, martinez1985} incorporated the nonlinearity while modelling passively mode-locked lasers. A linear iterative method was used for dispersion-tuned actively mode-locked lasers recently by Burgoyne \emph{et al.}~\cite{burgoyne2014}. Each of these models used a prescribed transfer function to describe the effect each component has on the pulse. While the transfer functions are chosen suitably to have the effect observed in the laboratory, the transfer functions are phenomenologically chosen, and not necessarily obtained directly from the underlying physics. Furthermore, these methods have not included the effect of every component, lacking the imperative interplay of dispersion, modulation, and nonlinearity---hindering their models to exhibit modulation instability.

@@

\subsection{Non-Dimensionalization}
Non-dimensionalization is a technique used in mathematical modelling to better understand the relative importance of different processes within a system. In essence, we scale all variables by typical values, based on the geometry and boundary conditions (see, for example,~\cite{howison2005}). Doing so `factors' the dimensional units out of the problem, leaving only dimensionless equations with dimensionless parameters. The relative magnitudes of the dimensionless parameters characterize the dominant dynamics---like how the Reynolds number can describe fluid flow. Each process in the laser can be better understood by re-scaling the time, energy, and amplitude by the convenient scalings:
\begin{align}
	T &= T_M \widetilde{T},& E &= \Es \widetilde{E},& A &= \left( \frac{\Es}{T_M} \right)^{1/2} \widetilde{A}.
\end{align}

@@

Now that we have the algebraic effect of each section of the cavity, we are ready to take the composition of the maps to give the effect of one round trip of the cavity. Unlike the models described in the beginning of this section, the permutation of our components is important as the maps \eqref{eq:effects} do not necessarily commute with each other. The geometry of a particular laser will dictate the order of the process maps. We assume the construction of the laser is as in Fig.~\ref{fig:cavity}, and provide the following justifications. The effect of the nonlinearity will be maximal after the pulse passes through the gain fibre---this is the region where the pulse is most energetic. Thus, we take the nonlinearity component to follow the gain. Similarly, we want the loss component to directly follow the nonlinearity in an attempt to minimize the effect the nonlinearity has on the pulse. This leaves the dispersion and modulation components. For now, let us assume the CFBG precedes the modulator. Finally, we assume the loss component is our starting point since this corresponds to the optical coupler, and thus, allows us to more easily compare to experiments. More precisely, let
\begin{align}
	\mathcal{L}(A) = F(G(M(D(L(A)))))
	\label{eq:order}
\end{align}
denote one complete round trip of the cavity. Since a ring laser is cyclic in nature, we use the result of one round trip as the input pulse for the next round trip---restarting the process. We continually repeat this procedure until the envelope and chirp of the pulse reach equilibrium, or the pulse succumbs to modulation instability. The phase change is uninteresting to us, as it is not modified by \eqref{eq:effects} in a meaningful way, and since the phase cannot easily be measured experimentally. An example of the evolution of the envelope during one round trip of the laser cavity can be found in Fig.~\ref{fig:cavityevo}.

@@

\subsection{Linear Solution}
By neglecting the effect of the nonlinearity, that is, $b = 0$, a solution can be found analytically. We proceed by assuming a fixed point solution will take the form of a linearly chirped Gaussian. The reasons for this assumption are that the solution to the models presented in~\cite{cutler1955, siegman1969, kuizenga1970a, martinez1984, martinez1985} were Gaussian, the equilibrium shape will correlate to the modulation function, and a Gaussian is a fixed point of the Fourier transform. Furthermore, we choose this form because it resembles the envelope and linear chirp expected from the literature~\cite{burgoyne2014, haus1975, haus1996, haus2000, usechak2005}.

@@

Consider the pulse
\begin{align}
	A = \sqrt{P} \exp \left( -(1 + iC) \frac{T^2}{2 \sigma^2} \right) \exp(i \phi_0),
	\label{eq:A0}
\end{align}
where $P$ is the peak power, $C$ is the chirp, $\sigma^2$ is the variance, and $\phi_0$ is the phase (Fig.~\ref{fig:samplegauss}). By computing \eqref{eq:A0} after one round trip, and solving $\mathcal{L}(A) = A \textrm{e}^{i \phi}$, a tedious manipulation yields the conditions
\begin{align}
	\frac{\sigma^4}{1 - \sigma^2} &= (\sigma^2 + 2 C s^2)^2 + 4 s^4, \label{eq:sigrelation} \\
	\frac{C}{1 - \sigma^2} &= C + 2 \frac{s^2}{\sigma^2} (1 + C^2), \label{eq:chirprelation} \\
	1 &= \frac{W(a E_g \textrm{e}^{E_g})}{E_g} h^2 (1 - \sigma^2)^{1/2} \label{eq:energyrelation}
\end{align}
for \eqref{eq:A0} to be in equilibrium. Manipulating \eqref{eq:sigrelation} and \eqref{eq:chirprelation} we are able to eliminate the chirp, and find the expression
\begin{align}
	\sigma^8 + 4 s^4 \sigma^6 - 20 s^4 \sigma^4 + 32 s^4 \sigma^2 - 16 s^4 = 0,
	\label{eq:var}
\end{align}
which contains $\sigma^2$ terms only. As \eqref{eq:var} is a quartic in $\sigma^2$, it has an analytic solution, namely,
\begin{equation}
	\begin{split}
		\sigma^2 = \sqrt{2} s \left( s^6 + 3s^2 + \sqrt{4 + s^4} \left( 1 + s^4 \right) \right)^{1/2} & \\
		- s^4 - s^2 & \sqrt{4 + s^4}.
		\label{eq:equilvar}
	\end{split}
\end{equation}
Additionally, the chirp is
\begin{align}
	C = \frac{\sigma^4 - \left( \sigma^8 - 16 s^4 \left( 1 - \sigma^2 \right)^2 \right)^{1/2}}{4 s^2 \left( 1 - \sigma^2 \right)},
	\label{eq:chirp}
\end{align}
in terms of $s$ and $\sigma^2$. Moreover, we can obtain the energy of the pulse entering the gain fibre, $E_g$, from \eqref{eq:energyrelation}, and thus, the equilibrium energy at the optical coupler. Doing so gives
\begin{align}
	E &= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{1 - h^2 \left( 1 - \sigma^2 \right)^{1/2}}.
	\label{eq:analenergy}
\end{align}
By equating the energy of the pulse in \eqref{eq:A0} and \eqref{eq:analenergy}, we find the peak power to be
\begin{align}
	P &= \frac{\log \left( a h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}{\sqrt{\pi} \sigma \left( 1 - h^2 \left( 1 - \sigma^2 \right)^{1/2} \right)}.
	\label{eq:analpower}
\end{align}

The nature of \eqref{eq:var}--\eqref{eq:analpower} makes the expressions difficult to work with. However, by asymptotically expanding \eqref{eq:var}--\eqref{eq:analpower} in terms of the dispersion parameter, $s$, we find the simpler relations
\begin{align}
	\sigma^2 &\sim 2s(1 - s) + \bigO{s^3}, \\
	C &\sim 1 - s + \frac{1}{2}s^2 + \bigO{s^3}, \\
	E &\sim \frac{\log (a h^2)}{1 - h^2} - \frac{1}{1 - h^2} \left( 1 + \frac{h^2 \log(a h^2)}{1 - h^2}  \right) s + \bigO{s^2}, \label{eq:maxenergy} \\
	P &\sim \frac{\log(ah^2)}{\sqrt{2 \pi}(1 - h^2)} s^{-1/2} + \bigO{s^{1/2}},
\end{align}
in the case $s \rightarrow 0$. And,
\begin{align}
	\sigma^2 &\sim 1 - \frac{1}{4}s^{-4} + \bigO{s^{-8}}, \label{eq:siginfty}\\
	C &\sim \frac{1}{2}s^{-2} + \bigO{s^{-6}},
\end{align}
in the case $s \rightarrow \infty$. We have no asymptotic expansion of the energy or peak power as $s \rightarrow \infty$ since in this limit the pulse loses too much energy due to dispersion and modulation. The gain is not powerful enough to balance the energy lost. The argument of the logarithm in \eqref{eq:analenergy} gives us a condition for the pulse to be sustainable. Expanding the requirement $a h^2 (1 - \sigma^2)^{1/2} > 1$, from \eqref{eq:analenergy}, using \eqref{eq:siginfty} (for simple approximation), we find 
\begin{align}
	s^* = \left( \frac{a h^2}{2} \right)^{1/2}
\end{align}
is the maximum value of $s$ to facilitate a sustainable pulse.

@@

In Fig.~\ref{fig:chirpft} we show the chirp and Fourier transform of the pulse \eqref{eq:nlA0} after it has reached equilibrium for the nominal parameter values given in \eqref{eq:ndparam}, and two additional $b$ values. In this nonlinear case, we find that the equilibrium pulse is no longer Gaussian---as evidenced by the Fourier transforms. Notice that compared to the linear case, the nonlinearity introduces higher frequency modes, giving a bi-modal distribution. Moreover, we find the pulse is linearly chirped near the peak, but, in the tails, $|T| > 1$, the chirp begins to saturate---consistent with the experimental results~\cite{chen2008, rothenberg1989b, tomlinson1985}. As the nonlinearity parameter, $b$, increases, the chirp increases more sharply across the pulse, and saturates at a larger value. Also, the two peaks in the Fourier transform become more pronounced and further apart, and the Fourier transform of the pulse becomes more rectangular. However, after a critical point, SPM plays a more substantial role, leading to modulation instability. Fig.~\ref{fig:breakevo} highlights this phenomenon. By decreasing the dispersion from to Fig.~\ref{fig:chirpft}, we find the pulse begins to `breathe'. During the first two dozen round trips of the cavity, the SPM compounds and becomes detrimental to the pulse. This in turn induces modulation instability, and degrades the pulse, until the pulse is no longer stable or sustainable. On the other hand, having a more moderate value of $b$, the pulse is able to equilibrate even with a less favourable, random, initial pulse (Fig.~\ref{fig:convevo}), as in~\cite{meng2020}. The pulse is able to shed the initial left lobe ($T \approx -1 / 2$) because of dispersion and modulation, and a central lobe ($T \approx 0$) forms and is able to grow due to the gain medium. This then allows the pulse to come to equilibrium very quickly. Furthermore, since the intensity of the initial pulse is relatively small, the effect of SPM is negligible during the first few round trips, allowing the laser to select preferable modes. The preferable modes then get amplified, stabilizing the pulse. It is clear the initial shape of the pulse is much less important than the initial energy of the pulse, $E_0$, or the strength of the nonlinearity, $b$.

@@

We now turn our attention to a more quantitative analysis, and characterize the stability of a laser by exploring the parameter space. We focus on the $s$--$b$ plane in particular, since $a$ and $h$ effect only the amplitude of the pulse, whereas the interplay of dispersion and nonlinearity gives rise to the instabilities. Additionally, as mentioned, a conservative initial energy will effect the dynamics much less than a more intense pulse since the nonlinearity will be negligible while initially coming to equilibrium. To identify the stability of a pulse, we examine the relative change in the pulse's envelope between consecutive round trips of the cavity. We compute this error as
\begin{align}
	\Delta = \frac{\| |\mathcal{L}^n(A_0)| - |\mathcal{L}^{n-1}(A_0)| \|_2}{\| \mathcal{L}^{n-1}(A_0) \|_2},
	\label{eq:error}
\end{align}
where
\begin{align}
	\| f \|_2^2 \coloneqq \int_{-\infty}^\infty |f|^2 \, \df T,
\end{align}
and we choose $n$ to be sufficiently large to guarantee the pulse reaches equilibrium or deteriorates due to modulation instability---we use $n = 500$ in our experiments. We take the modulus of the pulse since we are only interested in the evolution of the envelope because of the auto-correlation methods used for experimental measurements. We plot the error in Fig.~\ref{fig:error}. Two distinct regions split the parameter space, divided by an incredibly sharp and complicated boundary. In the upper-left region, the error is $\bigO{1}$; here, the nonlinearity is too strong, and the SPM induces modulation instability. In the lower-right region we find the opposite behaviour---the pulse reaches an equilibrium state and is stable. The increase in dispersion, coupled with modulation, allows the laser to balance the nonlinear effects. The error in this region is on the order of machine precision.

@@

Also of interest is the energy of the pulse at equilibrium, which we compute numerically by \eqref{eq:energy}. Given our ordering of the components, \eqref{eq:order}, the energy of the pulse is immediately before it passes through the optical coupler. The energy of each pulse emitted by the laser is $(1 - h^2) E$, and $h^2 E$ remains in the cavity. We plot the energy in the $s$--$b$ plane in Fig.~\ref{fig:energy}. Unsurprisingly, we find the same sharp boundary as we saw in Fig.~\ref{fig:error}. In the unstable region (upper-left) the energy is relatively small. Since the pulse is not sustainable in this region it is not able to foster prominent modes, thus, the intensity is low across the entire pulse. In the stable region (lower-right) we find that the energy smoothly decays as both $b$ and $s$ increase, with the contours being hyperbolic-esque. As $s$ increases, more energy is lost due to modulation since dispersion increases the length of the tails of the pulse. Our choice of modulation function suppresses the additional frequencies generated by SPM, as $b$ increases. We find that a more energetic laser requires weaker dispersion, however, the nonlinearity must be correspondingly low, otherwise modulation instability will destroy the pulse. Moreover, the maximum energy, which occurs in the limit that both $s$ and $b$ go to zero, is 2.5487. With our choice of parameters, \eqref{eq:maxenergy} predicts an equilibrium energy of 2.5535 at the origin, confirming the usefulness of the approximation.

@@

\section{Conclusion}
\label{sec:conclusion}
Expanding upon the ideas originally proposed by Cutler~\cite{cutler1955}, and Kuizenga and Siegman~\cite{kuizenga1970, kuizenga1970a, siegman1969}, we developed a nonlinear iterative map model for tuneable ring lasers. We recovered linearly chirped Gaussian solutions in the case of a Gaussian modulation function when omitting the nonlinearity. Moreover, the solution to the linear model extends to any modulation function using results of~\cite{calcaterra2008a}. In contrast, with the inclusion of the nonlinearity, we were able to recover wave breaking and modulation instability, and found a sharp boundary of stability. This phenomenon has been demonstrated in a laboratory setting~\cite{agrawal2013, anderson1992, finot2008, rothenberg1989b, tomlinson1985}, but, has proven difficult to predict with simple mathematical models~\cite{meng2020}. We have shown that a simple iterative map model is able to reproduce complicated instabilities arising naturally from the interplay between dispersion, modulation, and nonlinear effects. Our model easily adapts to specific laser geometries, and to more complicated models of the components. For example, incorporating a frequency dependence in the gain function, or using a saturable absorber instead of a Gaussian modulator.

Analyzing the nonlinear model analytically using an asymptotic expansion for small values of the nonlinearity parameter, $b$, is of interest. Doing so will provide insight to how the nonlinearity impacts the linear solution to better understand the manifestation of wave breaking and modulation instability. Furthermore, a more detailed analysis of crossing the boundary between stability and instability may provide valuable insight for mechanisms that trigger the instability.
